Weighted Model Counting Without Parameter Variables

### 2021-02-18 doc/paper3/paper.tex
The correctness proof is almost complete

\begin{theorem}[Early Projection
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20}, verbatim]
  Let $X$ and $Y$ be sets of variables. For all functions $f\colon 2^X \to
  \mathbb{R}$ and $g\colon 2^Y \to \mathbb{R}$, if $x \in X \setminus Y$, then
  $\exists_x (f \cdot g) = (\exists_x f) \cdot g$.
\end{theorem}
% TODO: mention that formulas, clauses, and models are all treated as sets

@@

% NOTE: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition}
  Let $(\phi, X_I, X_P, w)$ be an old WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer.
\end{proposition}
\begin{proof}
  The answer of \cref{eq:new_wmc} is
  \begin{align*}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset) &= \sum_{Y \subseteq X_I \cup X_P} \left(\prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)(Y) \\
&= \sum_{Y \subseteq X_I \cup X_P} \underbrace{\left(\prod_{c \in \phi} [c]_0^1\right)}_f(Y) \underbrace{\left(\prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)}_g(Y)
  \end{align*}
  by \cref{def:new_wmc,lemma:sum,def:operations}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given an old WMC instance $(\phi, X_I, X_P,
  w)$, returns a new WMC instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a family of literals
      $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$;
      \end{enumerate}
    \item or \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item there is no clause $c \in \phi$ with $\neg p \in c$,
      \item and for any pair of clauses $c, d \in \phi$ such that $c \ne d$, $p
        \in c$ and $p \in d$, $\mathtt{Vars}(c) \cap \mathtt{Vars}(d) = \{ p
        \}$.
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  % TODO: finish the proof
  First, let us assume that \cref{condition:d02} applies and consider the
  collective effect of the loop on
  \crefrange{line:foreach1start}{line:foreach1end} for: 1) clauses that have a
  positive parameter variable, 2) clauses that have a negative parameter
  variable (removed), 3) clauses that don't have parameter variables (left
  unchanged).
\end{proof}

\paragraph{Notes.}
\begin{itemize}
\item The family of literals could be empty.
\item This just says that $p$ is equivalent to a conjunction.
\item The first group of conditions applies to \texttt{d02}, while the second
  group applies to \texttt{bklm16}.
\item For \texttt{cd05} and \texttt{cd06}, condition $2bi$ should be replaced
  with $w(\neg p) = 1$.
\end{itemize}

@@

\begin{definition} \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y),
  \]
  and
  \[
    (fg)(Y) \coloneqq f(Y)g(Y)
  \]
  for all $Y \subseteq X$.
\end{definition}

\begin{definition} \label{def:restriction}
  Let $f\colon 2^X \to \mathbb{R}$ be a pseudo-Boolean function, and $x \in X$.
  Then $f|_{x=0}, f|_{x=1}\colon 2^X \to \mathbb{R}$ are \emph{restrictions} of
  $f$ defined as
  \[
    f|_{x=0}(Y) \coloneqq f(Y \setminus \{x\}),
  \]
  and
  \[
    f|_{x=1}(Y) \coloneqq f(Y \cup \{x\})
  \]
  for all $Y \subseteq X$.
\end{definition}

\begin{definition} \label{def:projection}
  Let $X$ be a set. For any $x \in X$, \emph{projection} $\exists_x$ is an
  endomorphism $\exists_x\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined
  as
  \[
    \exists_xf = f|_{x=1} + f|_{x=0}
  \]
  for any $f\colon 2^X \to \mathbb{R}$.
\end{definition}

@@

% NOTE: this is a rehash of equation (2) from the DPMC paper
\begin{definition} \label{def:abstraction}
  Let $(T, r, \gamma, \pi)$ be a PJT, and $X$ be the set of variables. The
  functionality of \textsc{DPMC} execution can be represented by $\delta(r)$,
  where $\delta\colon \mathcal{V}(T) \to \mathbb{R}^{2^X}$ is a recursive
  function defined as
  \begin{equation} \label{eq:execution}
    \delta(t) =
    \begin{cases}
      \gamma(t) & \text{if } t \in \mathcal{L}(T) \\
      \exists_{\pi(t)} \prod_{u \in \mathcal{C}(t)} \delta(u) & \text{otherwise.}
    \end{cases}
  \end{equation}
  The range of $\delta(r)$ then contains a single real number, i.e., the answer.
\end{definition}

>>>>

\begin{theorem}[Early Projection
  \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20},
  verbatim] \label{thm:early}
  Let $X$ and $Y$ be sets of variables. For all functions $f\colon 2^X \to
  \mathbb{R}$ and $g\colon 2^Y \to \mathbb{R}$, if $x \in X \setminus Y$, then
  $\exists_x (f \cdot g) = (\exists_x f) \cdot g$.
\end{theorem}
% TODO: mention that formulas, clauses, and models are all treated as sets

@@

% NOTE: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be an old WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer.
\end{proposition}
\begin{proof}
  The answer of \cref{eq:new_wmc} is
  \begin{align*}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset) &= \sum_{Y \subseteq X_I \cup X_P} \left(\prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)(Y) \\
&= \sum_{Y \subseteq X_I \cup X_P} \underbrace{\left(\prod_{c \in \phi} [c]_0^1\right)}_f(Y) \underbrace{\left(\prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)}_g(Y)
  \end{align*}
  by \cref{def:new_wmc,lemma:sum,def:operations}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given an old WMC instance $(\phi, X_I, X_P,
  w)$, returns a new WMC instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item and for any pair of clauses $c, d \in \phi$ such that $c \ne d$, $p
        \in c$ and $p \in d$, $\mathtt{Vars}(c) \cap \mathtt{Vars}(d) = \{ p
        \}$.
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem} % TODO: add parentheses around products
% TODO: maybe give a name to (2) since it's wrong to call it an equation
\begin{proof} % TODO: check if each condition is actually used
  % TODO: start here: consider how the transformation changes the 'normal form'.
  % We will show that every iteration keeps the answer the same as in the normal form.
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer as the given old WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I
        \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \cref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \cref{eq:new_wmc}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{condition:d02},
  \[
    \bigwedge_{c \in \phi_p} c \equiv p \Leftrightarrow \bigwedge_{i=1}^n l_i
  \]
  for some family of indicator variable literals $(l_i)_{i=1}^n$. Thus,
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 = \exists_p
    [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = \exists_p \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
      l_i \right]_0^1\right|_{p=0}
  \end{equation}
  by \cref{def:projection}. Since for any input, $\bigwedge_{i=1}^n l_i$ is
  either true or false, exactly one of the two summands in \cref{eq:bigsums}
  will be equal to one, and the other will be equal to zero, and so
  \[
    \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
      \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
        l_i \right]_0^1\right|_{p=0} = 1,
  \]
  where $1$ is a pseudo-Boolean function that always returns one. On the other
  side of \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged,
  we get $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied
  under \cref{condition:d02} when $w(p) = 1$.

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  By \cref{def:projection},
  \begin{align*}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 &= \left.\left( [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=1} + \left.\left([p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=0} \\
                                                                                      &= w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[ \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i \right]_1^{w(p)}
  \end{align*}
  as required. This finishes the proof of the correctness of the first `foreach'
  loop under \cref{condition:d02}.

  Now let us assume \cref{condition:bklm16}. We still want to prove
  \cref{eq:to_show}.
\end{proof}
% TODO: now consider the second for loop
% TODO: condition 1 is necessary in both cases because we're ignore the weights
% of indicator variables

\paragraph{Notes.}
\begin{itemize}
\item This just says that $p$ is equivalent to a conjunction.
\item The first group of conditions applies to \texttt{d02}, while the second
  group applies to \texttt{bklm16}.
\item For \texttt{cd05} and \texttt{cd06}, condition $2bi$ should be replaced
  with $w(\neg p) = 1$.
\end{itemize}

@@

\begin{definition} \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (fg)(Y) \coloneqq
    f(Y)g(Y)
  \]
  for all $Y \subseteq X$.
\end{definition}

\begin{definition} \label{def:restriction}
  Let $f\colon 2^X \to \mathbb{R}$ be a pseudo-Boolean function, and $x \in X$.
  Then $f|_{x=0}, f|_{x=1}\colon 2^X \to \mathbb{R}$ are \emph{restrictions} of
  $f$ defined as
  \[
    f|_{x=0}(Y) \coloneqq f(Y \setminus \{x\}), \quad \text{and} \quad
    f|_{x=1}(Y) \coloneqq f(Y \cup \{x\})
  \]
  for all $Y \subseteq X$.
\end{definition}

\begin{definition} \label{def:projection}
  Let $X$ be a set. For any $x \in X$, \emph{projection} $\exists_x$ is an
  endomorphism $\exists_x\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined
  as
  \[
    \exists_xf \coloneqq f|_{x=1} + f|_{x=0}
  \]
  for any $f\colon 2^X \to \mathbb{R}$.
\end{definition}

@@

% NOTE: this is a rehash of equation (2) from the DPMC paper
\begin{definition} \label{def:abstraction}
  Let $(T, r, \gamma, \pi)$ be a PJT, and $X$ be the set of variables. The
  functionality of \textsc{DPMC} execution can be represented by $\delta(r)$,
  where $\delta\colon \mathcal{V}(T) \to \mathbb{R}^{2^X}$ is a recursive
  function defined as
  \begin{equation} \label{eq:execution}
    \delta(t) \coloneqq
    \begin{cases}
      \gamma(t) & \text{if } t \in \mathcal{L}(T) \\
      \exists_{\pi(t)} \prod_{u \in \mathcal{C}(t)} \delta(u) & \text{otherwise.}
    \end{cases}
  \end{equation}
  The range of $\delta(r)$ then contains a single real number, i.e., the answer.
\end{definition}
